---
title: "Collection of plots"
subtitle: ""
author: "Airas Group"
institute: "RStudio, PBC"
date: "2025/5/9 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# Plots for exploratory analysis



---

## 1. Boxplots


```{r}
library(ggplot2)
library(ggpubr)
library(ggdist)  #for half-eye / raincloud plots

set.seed(123) #for the reproducibility of generated data
df <- data.frame(
  group = rep(c("Control", "Treatment"), each = 50),
  value = c(rnorm(50, mean=5, sd=1),
            rnorm(50, mean=6, sd=1.2))
)
```



---

### Classic boxplot

- Run the code:

```{r, echo=F}
library(ggplot2)
ggplot(df, aes(x = group, y = value, fill = group)) +
  geom_boxplot(alpha=0.9) +
  theme_light() +
  scale_fill_manual(values = c("Control" = "#1b9e77",
                               "Treatment" = "#d95f02"))+
  labs(title = "Boxplot by group", y = "Value", x = "Group")
```


---

### Boxplot with statistics

- Run the code:

```{r, echo=F}
library(ggplot2)
library(ggpubr)
# Compute p-value manually
p <- t.test(value ~ group, data = df)$p.value
p_label <- ifelse(p < 0.001, "<0.001", sprintf("%.3f", p))

ggplot(df, aes(x = group, y = value, fill = group)) +
  geom_boxplot() +
  annotate("text", x = 1.5, y = max(df$value) + 0.5, label = paste0("p: ", p_label), size = 5) +
  theme_light() +
   scale_fill_manual(values = c("Control" = "#1b9e77",
                               "Treatment" = "#d95f02"))+
  labs(title = "Boxplot with formatted p-value", y = "Value", x = "Group")

```

---

### Group-wise comaprison

- Run the code:

```{r, echo=F}

library(ggplot2)

set.seed(123)
df2 <- data.frame(
  group = rep(c("Control", "Treatment1", "Treatment2", "Treatment3"), each = 30),
  value = c(rnorm(30, 5, 1),
            rnorm(30, 6, 1.2),
            rnorm(30, 5.5, 1),
            rnorm(30, 6.5, 1.1))
)


# Boxplot with different colors
ggplot(df2, aes(x = group, y = value, fill = group)) +
  geom_boxplot() +
  stat_compare_means(method = "anova", label = "p.signif") +  # ANOVA for >2 groups
  scale_fill_manual(values = c("Control" = "#1b9e77",
                               "Treatment1" = "#d95f02",
                               "Treatment2" = "#7570b3",
                               "Treatment3" = "#e7298a")) +
  theme_light() +
  labs(title = "Boxplot for multiple comparison", y = "Value", x = "Group")

```

---

### Multiple comparisons

- Run the code:

```{r, echo=FALSE}
library(ggplot2)
library(ggpubr)

# Define all pairwise comparisons
my_comparisons <- list(
  c("Control", "Treatment1"),
  c("Control", "Treatment2"),
  c("Control", "Treatment3"),
  c("Treatment1", "Treatment2"),
  c("Treatment1", "Treatment3"),
  c("Treatment2", "Treatment3")
)

# Boxplot with all pairwise comparisons and p-values
ggplot(df2, aes(x = group, y = value, fill = group)) +
  geom_boxplot() +
  stat_compare_means(
    comparisons = my_comparisons, 
    method = "t.test", 
    label = "p.format",     # show exact p-values
    hide.ns = TRUE           # hide non-significant p-values if desired
  ) +
  scale_fill_manual(values = c("Control" = "#1b9e77",
                               "Treatment1" = "#d95f02",
                               "Treatment2" = "#7570b3",
                               "Treatment3" = "#e7298a")) +
  theme_light() +
  labs(title = "Boxplot with pairwise p-values", y = "Value", x = "Group")
```




---



## 2. Raincloud plots

We want to show boxplot, disperison of values and density of a sampled numeric variables, highlighitng the difference of distribution between two groups


---

- Run the code:

```{r, echo=F}
# Fancy raincloud plot
library(ggplot2)
library(ggdist) 

ggplot(df, aes(x=group, y=value, fill=group)) +
  ggdist::stat_halfeye(
    adjust = 0.5, width = 0.6, .width = 0, justification = -0.3
  ) +
  geom_boxplot(width=0.2, outlier.shape = NA, alpha=0.5) +
  geom_jitter(width=0.1, alpha=0.6) +
  scale_fill_brewer(palette="Set2") +
  labs(title="Raincloud plot",
       y="Measurement",
       x="Group") +
  theme_minimal(base_size = 14)
```


---



#  Plots for model estimates



---


## 1. Dot-and-whisker plot

Dot-and-whisker plots visually summarize regression results by showing the point estimate (dot) and confidence interval (whiskers) for each predictor. They are useful for quickly  comparing effect sizes and significance across multiple variables in a single, clean plot.

- We start from running logistic regression model 

- We obtain estimate (with p-values, confidence intervals) for each of the regressor in the model

```{r, echo=FALSE, message=F, warning=F}
# ---- Load libraries ----
library(dplyr)
library(ggplot2)
library(dotwhisker)
library(forcats)

# ---------------------- Example data ----------------------
 results_clean <- data.frame(
   Predictor = rep(c("Age", "Gender", "BMI", "HOT_PET"), each = 2),
   OR = c(1.25, 1.30, 0.85, 0.88, 1.10, 1.20, 1.40, 1.35),
   CI_low = c(1.10, 1.15, 0.65, 0.70, 0.90, 1.05, 1.10, 1.15),
  CI_high = c(1.40, 1.50, 1.05, 1.10, 1.30, 1.35, 1.70, 1.55),
  p_value = c(0.01, 0.02, 0.04, 0.03, 0.12, 0.10, 0.001, 0.002),
   model = rep(c("Unadjusted", "Adjusted"), 4)
 )
results_clean
```


---


- Now we need to arrange the data before plotting

- From the model output (logistic regression) we need to run this:
```{r}
dw_data <- results_clean %>%
  
  rename(
    estimate = OR,
    conf.low = CI_low,
    conf.high = CI_high,
    term = Predictor
  ) %>%
  mutate(
    model = factor(model, levels = c("Unadjusted", "Adjusted")),
    term = fct_rev(factor(term)),
    label = sprintf("%.2f", estimate),
    sig = ifelse(p_value < 0.05, "Significant", "Not significant")
  )
```

---

- And on the dw_data arranged for the plot, we run the dwplot function:
```{r, echo=F, warning=F}
dwplot(dw_data, 
       dot_args = list(size = 3), 
       whisker_args = list(size = 1),
       dodge_size = 0.6) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray40", size = 0.8) +
  scale_x_log10() +
  theme_minimal(base_size = 13) +
  labs(
    title = "Odds Ratios with 95% Confidence Intervals",
    subtitle = "Comparison of Unadjusted vs Adjusted Models",
    x = "Odds Ratio (log scale)",
    y = "",
    color = "Model"
  ) +
  scale_color_manual(values = c("Unadjusted" = "#1f77b4", "Adjusted" = "#ff7f0e")) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 12, color = "gray30"),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 11)
  ) +
  # Add OR labels near the dots
  geom_text(
    data = dw_data,
    aes(x = estimate, y = term, label = label, color = model),
    hjust = -0.5,
    size = 3.5,
    show.legend = FALSE
  ) +
  coord_cartesian(xlim = c(0.5, max(dw_data$conf.high) * 1.5)) +
  guides(color = guide_legend(override.aes = list(size = 4)))

```

---


# Plots for accuracy/ frequency tables



---


## 1. Mosaic plot 

Mosaic plots are used to visualize contingency tables. The size of each tile represents the proportion of observations in each group-category combination, allowing for quick comparisonof relative frequencies.

- We estart from a contingency table as the following one: 
```{r}
mat <- matrix(c(73, 27, 23, 77),
              nrow = 2,
              byrow = TRUE,
              dimnames = list(
                Group = c("PF%< 2.12", "PF% >= 2.12"),
                Category = c("Low", "High")
              ))
mat
```

---

- Next we run the mosaicplot function:

```{r, warning=F, echo=F}
mp <- mosaicplot(
  mat,
  color = c("#FF9800", "#03A9F4"),  # Orange & Blue
  main = "Distribution o by ventricular PF%",
  xlab = "Ventricular PF% group",
  ylab = "Phenotype",
  cex.axis = 1.2,
  cex.lab = 1.3,
  cex.main = 1.5,
  las = 1,       # horizontal labels on y-axis
  border = "white"
)

# Convert counts to percentages
total <- sum(mat)
percentages <- round((mat / total) * 100, 1)

# Get coordinates for annotations
# These values depend on the number of rows and columns.
x_coords <- rep(c(0.30, 0.75), times = nrow(mat))
y_coords <- c(0.70, 0.85, 0.15, 0.30)

# Flatten the matrix into a vector to match coordinates
percentage_labels <- paste0(as.vector(percentages), "%")

# Overlay text
text(
  x = x_coords,
  y = y_coords,
  labels = percentage_labels,
  cex = 1.2,
  font = 2,
  col = "black"
)

```

---

# Plots for longitudinal data



--- 


## 1. Spaghetti plot 

A spaghetti plot is a type of line plot used to visualize longitudinal or repeated-measures data for multiple subjects. The key idea is to plot each subject’s trajectory over time, so you can see both individual trends and the overall pattern.

- We start from a dataset for example with the following variables:  
```{r}
set.seed(42) # the seed allows to generate the same values for rBRLs
 df <- data.frame(
   ID = rep(1:20, each = 2),
   Group = rep(c("Treated", "Untreated"), each = 20),
   Time = rep(c("baseline", "follow-up"), times = 20),
   rBRL = c(rnorm(20, 1, 0.2), rnorm(20, 1.5, 0.2))
 )

```

---

- Since some values will coincide, we would like to randomly jitter them (spiking in the 2d-space)

- We will prepare the data before plotting:

```{r}
set.seed(42) # the seed allows to generate the same plot
df_jittered <- df %>%
  mutate(
    Time_num = as.numeric(factor(Time, levels = c("baseline", "follow-up"))),
    Time_jittered = Time_num + runif(n(), -0.08, 0.08) # jitter for clarity
  )
```


---

- Before plotting calculate the means:
```{r}
# Group means (without jitter)
group_means <- df %>%
  group_by(Group, Time) %>%
  summarise(mean_rBRL = mean(rBRL), .groups = "drop") %>%
  mutate(Time_num = as.numeric(factor(Time, levels = c("baseline", "follow-up"))))


```


---

- Run entirely the code 

```{r, echo=F}
ggplot() +
  # Individual subject lines and points
  geom_line(
    data = df_jittered,
    aes(x = Time_jittered, y = rBRL, group = ID, color = Group),
    alpha = 0.3, size = 0.8
  ) +
  geom_point(
    data = df_jittered,
    aes(x = Time_jittered, y = rBRL, color = Group),
    size = 2, alpha = 0.7
  ) +
  
  # Group-level mean lines and points (no ID mapping)
  geom_line(
    data = group_means,
    aes(x = Time_num, y = mean_rBRL, group = Group),
    color = "black", size = 1.2
  ) +
  geom_point(
    data = group_means,
    aes(x = Time_num, y = mean_rBRL),
    shape = 21, fill = "white", color = "black", size = 3, stroke = 1
  ) +
  
  # Axis and colors
  scale_x_continuous(
    breaks = c(1, 2),
    labels = c("Baseline", "Follow-up")
  ) +
  scale_color_manual(values = c("Treated" = "#1f77b4", "Untreated" = "#ff7f0e")) +
  
  # Titles and theme
  labs(
    title = "Longitudinal change in rBRL by group",
    x = "Time",
    y = "rBRL Value",
    color = "Group"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "top"
  )

```



---