---
title: "Correlation analysis"
author: "Joel Tuomaala"
date: "2025-09-03"
output: html_document
---
## Introduction

This workbook demonstrates how to perform correlation analysis between sets of numeric variables, optionally grouped by a categorical variable.  

The `correlation_analysis` function calculates correlation coefficients, confidence intervals, and p-values, using methods such as **Spearman**, **Pearson**, or **Kendall**.  

### When to use

- You have continuous numeric variables and want to assess pairwise associations.  
- You may have **groups** (e.g., treatment vs control) and want group-specific correlations.  
- You need **confidence intervals** for the correlation coefficients.

---

### Data Preparation

```{r, warning=F}
# Load required packages
library(dplyr)
library(psych)
library(knitr)

# Example: mtcars dataset
data <- mtcars

# Define variables
dependent <- "mpg"
independent <- c("wt", "hp")
group_column <- "am" # automatic/manual transmission

# Inspect data
head(data)
summary(data[ , c(dependent, independent, group_column)])
```

```{r}

# Defining the function
correlation_analysis <- function(
    input_data = mtcars,                          
    dependent = "mpg",                            
    independent = c("wt", "hp"),                  
    method = "spearman",                          
    group_column = "am"                           
) {
  
  # Load required packages
  if(!require(dplyr)) install.packages("dplyr")
  if(!require(psych)) install.packages("psych")
  if(!require(rlang)) install.packages("rlang")
  library(dplyr)
  library(psych)
  library(rlang)
  
  # Input validation
  if (!is.data.frame(input_data)) stop("input_data must be a data frame")
  if (!all(dependent %in% colnames(input_data))) stop("Not all dependent variables exist")
  if (!all(independent %in% colnames(input_data))) stop("Not all independent variables exist")
  if (!method %in% c("pearson","spearman","kendall")) stop("Invalid method")
  if (!group_column %in% colnames(input_data)) stop("Group column does not exist")
  
  correlation_results <- data.frame(
    Variable1 = character(),
    Variable2 = character(),
    CorrelationCoefficient = numeric(),
    RawLower = numeric(),
    RawUpper = numeric(),
    P_value = numeric(),
    AdjLower = numeric(),
    AdjUpper = numeric(),
    Group = character(),
    stringsAsFactors = FALSE
  )
  
  # Check numeric
  for (var in c(dependent, independent)) {
    if (!is.numeric(input_data[[var]])) stop(paste("Variable", var, "is not numeric"))
  }
  
  # Loop over groups
  for (group_loop in unique(input_data[[group_column]])) {
    collection <- input_data %>% filter(!!sym(group_column) == group_loop)
    if (nrow(collection) < 3) {
      warning(paste("Group", group_loop, "has fewer than 3 observations. Skipping."))
      next
    }
    
    correlation_score <- psych::corr.test(collection[independent], collection[dependent], method = method)
    
    for (i in seq_along(independent)) {
      for (j in seq_along(dependent)) {
        correlation_results <- rbind(correlation_results, data.frame(
          Variable1 = independent[i],
          Variable2 = dependent[j],
          CorrelationCoefficient = correlation_score$r[i,j],
          RawLower = correlation_score$ci$lower[i+(i*(j-1))],
          RawUpper = correlation_score$ci$upper[i+(i*(j-1))],
          P_value = correlation_score$p[i,j],
          AdjLower = correlation_score$ci.adj$lower[i+(i*(j-1))],
          AdjUpper = correlation_score$ci.adj$upper[i+(i*(j-1))],
          Group = as.character(group_loop),
          stringsAsFactors = FALSE
        ))
      }
    }
  }
  
  correlation_results$Significance <- NA
  correlation_results$Significance[correlation_results$P_value >= 0.05] <- "ns"
  correlation_results$Significance[correlation_results$P_value < 0.05] <- "*"
  correlation_results$Significance[correlation_results$P_value < 0.01] <- "**"
  correlation_results$Significance[correlation_results$P_value < 0.001] <- "***"
  
  return(correlation_results)
}
```



Assumption Checks

### 1. Check numeric variables

```{r}

sapply(c(dependent, independent), function(x) is.numeric(data[[x]]))
```

All variables must be numeric for correlation.

### 2. Check group sizes (if using grouping)

```{r}
table(data[[group_column]])
```


Each group should have at least 3 observations. Groups smaller than this will be skipped.

### 3. Explore relationships visually

```{r}
library(ggplot2)
```


```{r}

# Scatter plots of dependent vs each independent
for (var in independent) {
  ggplot(data, aes_string(x = var, y = dependent, color = group_column)) +
    geom_point(size = 2) +
    geom_smooth(method = "lm", se = FALSE) +
    theme_minimal() +
    ggtitle(paste("Scatter plot:", dependent, "vs", var))
}

```

Visual inspection helps detect non-linear relationships, outliers, or unusual patterns.

### Run Correlation Analysis

```{r}

# Run the correlation_analysis function
corr_results <- correlation_analysis(
  input_data = data,
  dependent = dependent,
  independent = independent,
  method = "spearman",
  group_column = group_column
)

```
### Showing results
```{r}
knitr::kable(head(corr_results), digits = 3, caption = "Correlation Analysis Results")

```


