---
title: "Univariate Logistic Regression (Cross-sectional)"
author: "Giulia Nicole Baldrighi"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

Univariate logistic regression is used to explore associations between predictors and a binary outcome in cross-sectional data.  

Before interpreting results, it is important to check model assumptions and fit:

1. Linearity of logit for continuous predictors  
2. No highly influential observations  
3. No extreme multicollinearity  
4. Goodness-of-fit

---

### Data Preparation


```{r, warning=F}
# Load required packages
library(dplyr)
library(broom)
library(ggplot2)
library(knitr)

# Example dataset
set.seed(123)
data <- data.frame(
  ID = 1:60,  
  Age = rnorm(60, 50, 10),
  BMI = rnorm(60, 25, 4),
  HOT_PET = sample(c("Yes", "No"), 60, replace = TRUE),
  Outcome = sample(c("Yes", "No"), 60, replace = TRUE)
)

# Ensure outcome is numeric 0/1
data$Outcome <- as.numeric(data$Outcome == "Yes")

# Select predictors
predictors <- c("Age", "BMI", "HOT_PET")
```


### Check frequences of the outcome (balanced categories)
Avoid categories with too low frequencies
```{r}
table(data$Outcome)
prop.table(table(data$Outcome))

```

### Check missing data
```{r}
sapply(data[predictors], function(x) sum(is.na(x)))

```


### Linearity

```{r}
# Add logit of outcome probability vs continuous predictors
for (var in c("Age", "BMI")) {
  model <- glm(as.formula(paste("Outcome ~", var)), data = data, family = binomial)
  plot(data[[var]], log(model$fitted.values / (1 - model$fitted.values)),
       xlab = var, ylab = "Logit of fitted probability",
       main = paste("Linearity check for", var))
  abline(lm(log(model$fitted.values / (1 - model$fitted.values)) ~ data[[var]]), col = "red")
}

```

The logit should show approximately linear relationship with continuous predictors.
## Influential observations 

```{r}
# Cook's distance for each predictor
for (var in predictors) {
  model <- glm(as.formula(paste("Outcome ~", var)), data = data, family = binomial)
  cook <- cooks.distance(model)
  plot(cook, type = "h", main = paste("Cook's distance for", var),
       ylab = "Cook's distance")
  abline(h = 4/(nrow(data)), col = "red", lty = 2)
}
```

Points above the red line may be highly influential and should be checked.


### Multicollinearity

```{r}
# Only relevant for continuous predictors
model_all <- glm(Outcome ~ Age + BMI, data = data, family = binomial)
car::vif(model_all)
```

VIF > 5–10 indicates high multicollinearity.


### Goodness of fit
```{r}
# Hosmer-Lemeshow test
if(!require(ResourceSelection)) install.packages("ResourceSelection")
library(ResourceSelection)

model <- glm(Outcome ~ Age + BMI + HOT_PET, data = data, family = binomial)
hoslem.test(data$Outcome, fitted(model))

```

p > 0.05 suggests good fit.

### Run univariate logistic regression

```{r}
results_list <- list()

for (var in predictors) {
  formula <- as.formula(paste("Outcome ~", var))
  model <- glm(formula, data = data, family = binomial)
  
  tidy_model <- broom::tidy(model) %>%
    filter(term == var) %>%
    mutate(
      OR = exp(estimate),
      CI_low = exp(estimate - 1.96 * std.error),
      CI_high = exp(estimate + 1.96 * std.error),
      p_value = p.value
    ) %>%
    select(term, OR, CI_low, CI_high, p_value)
  
  results_list[[var]] <- tidy_model
}

results_table <- bind_rows(results_list) %>%
  rename(Predictor = term)

knitr::kable(results_table, digits = 3, caption = "Univariate Logistic Regression Results (Cross-sectional)")
```


