---
title: "Univariate logistic regression with repeated measures"
author: "Giulia Nicole Baldrighi"
date: "2025-09-03"
output: html_document
---
## Introduction

Univariate logistic regression for repeated measures is useful when you want to test the association between individual predictor variables and a binary outcome while accounting for repeated measurements within subjects.  

This approach should be used when:

- outcome is binary (e.g., Yes/No, 0/1).  
- have repeated observations for the same subject over time.  
- the aim is to assess each predictor individually before considering multivariate models.  

We use a mixed-effects logistic regression model (`glmer` from `lme4`) to include a random intercept for each subject (`ID`) to account for within-subject correlation.

---

### Data Preparation

```{r}
# Load required packages
library(lme4)
library(dplyr)
```


### Run the models 

```{r}

set.seed(123)
data <- data.frame(
  ID = rep(1:20, each = 3),      # 20 subjects, 3 timepoints each
  Time = rep(1:3, times = 20),   # timepoints
  Age = rnorm(60, 50, 10),
  BMI = rnorm(60, 25, 4),
  HOT_PET = sample(c("Yes", "No"), 60, replace = TRUE),
  Outcome = sample(c("Yes", "No"), 60, replace = TRUE)
)

# Ensure outcome is numeric 0/1
data$Outcome <- as.numeric(data$Outcome == "Yes")

# Select predictors
predictors <- c("Age", "BMI", "HOT_PET")

```



```{r}
# Initialize empty results table
# Initialize empty list for results
results_list <- list()

for (var in predictors) {
  formula <- as.formula(paste("Outcome ~ Time + (1 | ID) +", var))
  model <- glmer(formula, data = data, family = binomial)
  
  # Tidy the model using broom.mixed
  tidy_model <- broom.mixed::tidy(model, effects = "fixed")
  
  # Select row corresponding to the predictor
  predictor_row <- tidy_model %>%
    filter(term == var) %>%
    mutate(
      OR = exp(estimate),
      CI_low = exp(estimate - 1.96 * std.error),
      CI_high = exp(estimate + 1.96 * std.error),
      p_value = p.value
    ) %>%
    select(term, OR, CI_low, CI_high, p_value)
  
  results_list[[var]] <- predictor_row
}

# Combine all results into a single table
results_table <- bind_rows(results_list) %>%
  rename(Predictor = term)

# Display results as a table
knitr::kable(results_table, digits = 3, caption = "Univariate Logistic Regression Results (Repeated Measures)")

```


