---
title: "Pre-post analysis"
author: "Giulia Baldrighi"
date: "2025-10-12"
output: html_document
---
  
## Delta-change

**Delta change** represents the **difference between post-treatment and baseline measurements**:

\[
\Delta = \text{Post} - \text{Pre}
\]

- Allows to see **how much a variable changes** after intervention.  
- Often used when comparing **treatment effects**.  

```{r}
# Example dataset
set.seed(123)
df <- data.frame(
  group = rep(c("Control", "Treatment"), each = 30),
  pre = c(rnorm(30, 50, 5), rnorm(30, 52, 5)),
  post = c(rnorm(30, 51, 5), rnorm(30, 57, 5))
)

# Compute delta
df$delta <- df$post - df$pre
head(df)


# Check normality of delta
shapiro.test(df$delta)  # Shapiro-Wilk test

# if normal: t-test
t.test(delta ~ group, data = df)

# if not: wilcox-test
wilcox.test(delta ~ group, data = df)
```


---

## Pre-Post paired analysis

* If the **same subjects** are measured before and after, use **paired tests**.

```{r}
# Paired t-test for Treatment group
t.test(df$post[df$group=="Treatment"], 
       df$pre[df$group=="Treatment"], 
       paired = TRUE)
```

* **Interpretation:** tests whether the **mean change from baseline** is statistically significant.

* **Alternative (non-parametric):** Mann-Withney signed-rank test if normality assumption is violated:

```{r}
wilcox.test(df$post[df$group=="Treatment"], 
            df$pre[df$group=="Treatment"], 
            paired = TRUE)
```

---

## Comparing delta across groups

* You can compare **delta values between groups** using t-test or ANOVA:

```{r}
# t-test for delta between Control and Treatment
t.test(delta ~ group, data = df)

# Or ANOVA for multiple groups
# (here just two groups)
anova_model <- aov(delta ~ group, data = df)
summary(anova_model)
```

* **Interpretation:** Tests whether **mean change differs** between groups.

---

## ANCOVA (Analysis of Covariance)

* ANCOVA adjusts **post-treatment outcomes** for baseline differences:

[
\text{Post} = \beta_0 + \beta_1 \text{Group} + \beta_2 \text{Pre} + \epsilon
]

```{r}
# ANCOVA adjusting post for pre
ancova_model <- lm(post ~ group + pre, data = df)
summary(ancova_model)
```

**Interpretation:**

- `group` effect tells if **treatment affects post**, **adjusted for baseline**.

- More powerful than delta alone if baseline differs across groups.

* Diagnostic checks:

```{r}
par(mfrow=c(2,2))
plot(ancova_model)
```

---

## Take home message

- **Delta analysis**: simple, intuitive, may ignore baseline differences.

- **Pre-post paired test**: good for within-subject changes.

- **ANCOVA**: adjusts for baseline, increases statistical power, handles covariates.











