---
title: "Assumptions and linear models with diagnostics"
author: "Giulia Nicole Baldrighi"
date: "2025-09-03"
output: html_document
---

## Introduction

Linear regression is used to model the relationship between a **continuous outcome** and one or more predictor variables.  

Before interpreting results, you should check the following assumptions:

1. **Linearity**: Relationship between predictors and outcome is linear.  
2. **Normality of residuals**: Residuals are normally distributed.  
3. **Homoscedasticity**: Constant variance of residuals across predicted values.  
4. **Independence of errors**.  
5. **No multicollinearity**: Predictors are not highly correlated.  
6. **No influential points**: No single observation unduly affects the model.

---

## Data Preparation

```{r, warning=F}
library(dplyr)
library(ggplot2)
library(car)
library(broom)
library(knitr)
library(reshape2)
```


```{r}
# Example dataset
set.seed(123)
data <- data.frame(
  ID = 1:60,
  Age = rnorm(60, 50, 10),
  BMI = rnorm(60, 25, 4),
  Exercise = rnorm(60, 3, 1),
  BloodPressure = rnorm(60, 120, 15)  # Continuous outcome
)

predictors <- c("Age", "BMI", "Exercise")
outcome <- "BloodPressure"
```


## Assumptions before the model

### Check distributions and outliers

```{r}
# Histogram and boxplot
data %>%
  select(all_of(c(outcome, predictors))) %>%
  melt(id.vars = NULL) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 15, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()

```



### Normality outcome 

```{r}
shapiro.test(data[[outcome]])
```
p > 0.05 suggests approximate normality.


### Homogeneity of variance (Levene’s test)


```{r}
library(car)
# Example: grouping by median split of Age
data$AgeGroup <- ifelse(data$Age > median(data$Age), "High", "Low")
leveneTest(data[[outcome]] ~ data$AgeGroup)
```

p > 0.05 suggests equal variance across groups.



### Correlation between predictors (multicollinearity)

```{r}
cor(data %>% select(all_of(predictors)))
vif(lm(as.formula(paste(outcome, "~", paste(predictors, collapse = "+"))), data=data))

```

VIF > 5–10 indicates high multicollinearity.


## Fit linear model
```{r}
results_list <- list()

for (var in predictors) {
  formula <- as.formula(paste(outcome, "~", var))
  model <- lm(formula, data = data)
  
  # Model diagnostics
  par(mfrow=c(2,2))
  plot(model, main = paste("Diagnostics for", var))
  
  tidy_model <- broom::tidy(model)
  results_list[[var]] <- tidy_model
}

results_table <- bind_rows(results_list) %>%
  rename(Predictor = term)

knitr::kable(results_table, digits = 3, caption = "Linear Regression Results for Each Predictor")

```

